---
title: 'SI - nonstructural carbon reserves'
author: Tim Rademacher
date: "22/03/2019"
output: html_document
bibliography: /Volumes/added_SSD/projects/NSF-DB_plant_growth/bib/Exp2017.bib
csl: /Volumes/added_SSD/projects/NSF-DB_plant_growth/bib/harvard.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library ('tidyverse')
```

## Calculating changes in nonstructural carbon reserves of stem section for the 2017 experiment at Harvard Forest

To estimate changes in soluble sugars and starch for various tissue, we scaled the concentrations derived from each tissue using colorimetric assays. Below we describe the scaling assumption for each tissue, which gave us total reserves in each organ. The differences of total values between the four sampling points in turn provide estimates of the gains or losses of soluble sugars and starch for each tissue and tree. To scale concentrations, the biomass for each tissue is first estimated using allometric equations from @jenkins_comprehensive_2004. We then multiply the tissue biomass by the nonstructural carbon concentration measured for that tissue similar to @furze_whole-tree_2019. Below we provide examples for each tissue.

## Tree and organ biomass

According to @jenkins_comprehensive_2004, we calcualted total aboveground biomass ($B; in \; kg$) from the diameter at breast height ($DBH; in \; m$) as follows:

$$
B = exp (\beta_0 + \beta_1 \; ln \; DBH)
$$

``` {r totalBiomass}
# Read tree stats (in particular dbh)
trees <- read_csv ('../../data/trees/2017ExperimentalDataPredictors.csv', col_types = cols ())

# Determine the total above ground biomass of each tree (biomass; kg)
biomass <- exp (2.5356 + 2.4349 * log (trees$dbh150)) 
```

with generic pine species (Pinus spp.) parameters of $\beta_0 = -2.5356$ and $\beta_1 = 2.4349$. This total biomass does include foliage, stemwood, bark and branches. To get tissue specific values for stemwood, bark and foliage we estimated the fraction of the total biomass in each tissue using the following equation:

$$
f = exp (\alpha_0 + \frac {\alpha_1} {DBH})
$$
$\alpha_0$ and $\alpha_1$ are wood type specific coeffecients. Here, we used the softwood values, for example $\alpha_0 = –0.3737$ and $\alpha_1 = –1.8055$ for stemwood, from Table 2 in @jenkins_comprehensive_2004.

``` {r foliageBiomass}
# Determine the fractions of total above ground biomass in foliage 
foliFraction <- exp (-2.9584 + ( 4.4766 / trees$dbh150))
rootFraction <- exp (-1.5619 + ( 0.6614 / trees$dbh150))
barkFraction <- exp (-2.0980 + (-1.1432 / trees$dbh150))
stemFraction <- exp (-0.3737 + (-1.8055 / trees$dbh150))
branFraction <- 1.0 - foliFraction - barkFraction - stemFraction

# Determine actual biomass in each organ
foliage  <- foliFraction * biomass
branches <- branFraction * biomass
roots    <- foliFraction * biomass
bark     <- foliFraction * biomass
stem     <- foliFraction * biomass
```

Branch biomass was estimated as the residual fraction of total biomass and the sum of foliage, bark and stem fractions.

## For the whole stem

To estimate the whole stem nonstructural carbon reserves, the biomass of the stem, as derived above, is further divided into the biomass of a hollow cylinder of the first centimeter and the inner cylinder. Each division is then multiplied with the respective nonstrucutral carbon concentration. 

## For stem sections

We measured nonstructural carbon concentrations for the first of the stem core under the bark and scaled to the entire section using a linear relationship, as concentrations gradients have been shown to vary among species but generally decline with distance from the bark [@hoch_non-structural_2003]. In particular, Pinus sylvestris - the closest relative to Pinus strobus - has been shown to have linearly declining stem wood nonstructrual carbon concentrations [@hoch_non-structural_2003].
<!-- The decline in @hoch_non-structural_2003 is actually more exponential than linear, but let's start with linear. Morgan might have better data for Pinus strobus that can help me scale more adequately.--> 

Assuming linear scaling from the measured nonstructural carbon concentration in the first centimeter under the bark to zero at the pith, we obtain nonstructural carbon concentrations across the radial profile for each tree. These concentrations can then be simply integrated across the stem section by multiplication of the mean concentration across the radial profile ($\bar c_{NSC}$) with the mass of the stem section ($M_{s}$) to obtain the total nonstructural carbon reserves in the stem section ($R_{NSC}$).

$$
R_{NSC} = \bar c_{NSC} \times M_{s}
$$

The mass of the stem section is derived from its volume ($V_{s}$) and density ($\rho_{s}$). For the density 289.9$\pm$40.4 $kg \, m^{-3}$ is used, which is the mean $\pm$ standard deviation of wood density of 16 cores from nine control white pine trees in the control group as determined using x-ray film. The volume is assumed to be a cylinder with diameter $dbh$ and height $h$. <!-- Maybe Patrick can give a reference for the exact procedure?-->
$$
M_{s} = V_{s} \times \rho_{s} = (\frac {d_s}{2})^2 \pi \times h \times \rho_{s}
$$

``` {r calculateStemSectionNSCs}
# Declare the height of the stem section (h, m)
h <- 0.1

# Declare mean stem wood density (rho_s, kg m-3)
rhos <- 289.9

# Read tree stats (in particular dbh)
trees <- read_csv ('../../data/trees/2017ExperimentalDataPredictors.csv', col_types = cols ())

# Determine the volume of each stem section (Vsection; m3)
trees$Vsection <- ((trees$dbh150 / 100.0) / 2.0)**2.0 * pi * h # dbh is in cm, therefore divide by 100.

# Determine the mass of each stem section (Msection, kg)
trees$Msection <- trees$Vsection * rhos

# Read the mean wood concentration in the first centimeter for each section
# TTR Do this once I have them.

# Get mean concentration across the radial profile (gC gDW-1) by division by two.
cSugar  <- 5.0 # this needs to be read in once I have values.
cStarch <- 3.0 # this needs to be read in once I have values.

# Get the total sugar, strach and NSC reserves in the stem section
RSugar <- trees$Msection * cSugar
RStarch <- trees$Msection * cStarch
RNSC <- RSugar + RStarch
```

## For needles

Similar to the stem's nonstructural carbon reserves, foliar biomass is multiplied with the mean foliar concentration of nonstructural carbon to obtain the total foliar reservoir of nonstructural carbon. The foliar biomass was derived as a fraction of the total above ground biomass according to @jenkins_comprehensive_2004.

## For roots

Finally root nonstructural carbon reserves were also scaled by multiplying the sum of coarse and fine root biomass with the root nonstrucutral carbon concentration. To estimate total root biomass, we divided the coarse root fraction as per @jenkins_comprehensive_2004 into three combinations of large and intermediate roots (90% and 10%, 75% and 25%, and 50% and 50%) and added a total fine root biomass equivalent to 5%, 15% and 30% of the coarse roots, respectively [@richardson_distribution_2015]. These are realistic scenarios used by @richardson_distribution_2015 to bracket the uncertainty of root biomass in white pine. <!-- Morgan did this differently! Maybe I should change this.-->

### To-do list and comments

- Get better data for radial variation in NSCs in white pine stems from Morgan, as the scaling from Pinus sylvestris after the data from Hoch et al. (2003) is not really adequate.
- adjust for more accurate dbh values once I measured the circumference for each section. At the moment is only calulates the mass once per tree and assume all sections have the same mass.

# References