---
title: 'SI - nonstructural carbon reserves'
author: Tim Rademacher
date: "06/09/2019"
output: pdf_document
bibliography: "/home/trademacehr/projects/NSF-DB\ Plant\ Growth/bib/Exp2017.bib"
csl: "/home/trademacehr/projects/NSF-DB\ Plant\ Growth/bib/harvard.csl"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library ('tidyverse')
library ('googlesheets')
```

## Calculating changes in nonstructural carbon reserves of stem section for the 2017 experiment at Harvard Forest

To estimate changes in soluble sugars and starch for various tissue, we scaled the concentrations derived from each tissue using colorimetric assays. Below we describe the scaling assumption for each tissue, which gave us total reserves in each organ. The differences of total values between the four sampling points in turn provide estimates of the gains or losses of soluble sugars and starch for each tissue and tree. To scale concentrations, the biomass for each tissue is first estimated using allometric equations from @jenkins_comprehensive_2004. We then multiply the tissue biomass by the nonstructural carbon concentration measured for that tissue similar to @furze_whole-tree_2019. Below we provide examples for each tissue.

## Tree and organ biomass

According to @jenkins_comprehensive_2004, we calcualted total aboveground biomass ($B; in \; kg$) from the diameter at breast height ($DBH; in \; m$) as follows:

$$
B = exp (\beta_0 + \beta_1 \; ln \; DBH)
$$

``` {r totalBiomass}
# Read tree stats (in particular dbh)
trees <- read_csv ('../../data/trees/2017ExperimentalDataPredictors.csv', col_types = cols ())

# Determine the total above ground biomass of each tree (biomass; kg)
biomass <- exp (2.5356 + 2.4349 * log (trees$dbh150)) 
```

with generic pine species (Pinus spp.) parameters of $\beta_0 = -2.5356$ and $\beta_1 = 2.4349$. This total biomass does include foliage, stemwood, bark and branches. To get tissue specific values for stemwood, bark and foliage we estimated the fraction of the total biomass in each tissue using the following equation:

$$
f = exp (\alpha_0 + \frac {\alpha_1} {DBH})
$$
$\alpha_0$ and $\alpha_1$ are wood type specific coeffecients. Here, we used the softwood values, for example $\alpha_0 = –0.3737$ and $\alpha_1 = –1.8055$ for stemwood, from Table 2 in @jenkins_comprehensive_2004.

``` {r foliageBiomass}
# Determine the fractions of total above ground biomass in foliage 
foliFraction <- exp (-2.9584 + ( 4.4766 / trees$dbh150))
rootFraction <- exp (-1.5619 + ( 0.6614 / trees$dbh150))
barkFraction <- exp (-2.0980 + (-1.1432 / trees$dbh150))
stemFraction <- exp (-0.3737 + (-1.8055 / trees$dbh150))
branFraction <- 1.0 - foliFraction - barkFraction - stemFraction

# Determine actual biomass in each organ
foliage  <- foliFraction * biomass
branches <- branFraction * biomass
roots    <- foliFraction * biomass
bark     <- foliFraction * biomass
stem     <- foliFraction * biomass
```

Branch biomass was estimated as the residual fraction of total biomass and the sum of foliage, bark and stem fractions.

## For the whole stem

To estimate the whole stem nonstructural carbon reserves, the biomass of the stem, as derived above, is further divided into the biomass of a hollow cylinder of the first centimeter and the inner cylinder. Each division is then multiplied with the respective nonstrucutral carbon concentration. 

## For stem sections

We measured nonstructural carbon concentrations for the first centimeter of the stem core (from the cambium towards the pith) as detailed in the methods section. While nonstructural carbon concentrations follow a generally decreasing gradient from the bark towards the pith [@hoch_non-structural_2003;furze_whole-tree_2018], we only consider nonstructural carbon reserves in the first centimeter underlying the bark for each stem section. We limit our estimation to the first centimeter, because deeper reserves are more difficult, and thus less likely, to be accessed on timescales considered here.
<!-- The decline in @hoch_non-structural_2003 is actually more exponential than linear, but let's start with linear. Morgan might have better data for Pinus strobus that can help me scale more adequately.--> 

Assuming that the bulk of easily accessible nonstructural carbon can be found in the first centimeter and that our measurements are an average of these reserves, we derive the total amount of nonstructural carbon in sugars and strach ($M_{sugar,i}$ and $M_{starch,i}$, respectively) by multiplying the mass of the hollow cylinder ($M_{s,i}$) of each stem section $i$ with the section's measured concentrations of sugar and starch ($c_{sugar,i}$ and $c_{starch,i}$, respectively). 

$$
 M_{sugar,i} = c_{sugar,i} \times M_{s,i} = V_{s,i} \times \rho_w \times c_{sugar,i}
$$
The mass of a hollow cylinder representing the outeter most centimeter of material is derived from its volume ($V_{s,i}$) and an average density of the wood of ... ($\rho_w$). In turn the volume calculated using a height ($h$) of 10 cm and the section's circumference ($cbh_{s,i}$; in $cm$), which was measured in the field using a tape measure. We use  289.9$\pm$ 40.4$kg \, m^{-3}$ as density, which is the mean $\pm$ standard deviation of wood density of 16 cores from nine trees in the control group as determined using x-ray film. <!--Maybe Patrick can giev a reference for the exact  procedure?-->

The mass of the stem section is derived from its volume ($V_{s}$) and density ($\rho_{s}$). For the density 289.9$\pm$40.4 $kg \, m^{-3}$ is used, which is the mean $\pm$ standard deviation of wood density of 16 cores from nine control white pine trees in the control group as determined using x-ray film. The volume is assumed to be a cylinder with diameter $dbh$ and height $h$. <!-- Maybe Patrick can give a reference for the exact procedure?-->
$$
M_{s} = V_{s} \times \rho_{s} = (\frac {d_s}{2})^2 \pi \times h \times \rho_{s}
$$

``` {r calculateStemSectionNSCs, warning = FALSE}
# Get the tree measurements of circumference from the googlesheet 
key <- '1K8Vk9VHRdwNjIa28frc-Zt4uyk07G6CZe1EkJunUll0'
suppressMessages (predictors <- gs_read (gs_key (key), ws = 'predictors', col_types = cols ()))

# Declare the height of the stem section (h, m)
h <- 0.1

# Declare mean stem wood density (rho_s, kg m-3)
rho_w <- 289.9

# Determine the volume of each stem section (m3) at various height (0.5m, 1m, 1.5m, 2m, 2.5m)
# cbh is in cm, therefore divide by 100 and by 2.0 and pi to get the sections radius in m
predictors <- add_column (predictors,   
                          V50  = ((predictors[['cbh50']]/(100.0*2.0*pi))^2.0*pi*h) - 
                                 ((predictors[['cbh50']]/(100.0*2.0*pi)-0.01)^2.0*pi*h),
                          V100 = (predictors[['cbh100']]/(100.0*2.0*pi))^2.0*pi*h - 
                                 ((predictors[['cbh100']]/(100.0*2.0*pi)-0.01)^2.0*pi*h),
                          V150 = (predictors[['cbh150']]/(100.0*2.0*pi))^2.0*pi*h - 
                                 ((predictors[['cbh150']]/(100.0*2.0*pi)-0.01)^2.0*pi*h), 
                          V200 = (predictors[['cbh200']]/(100.0*2.0*pi))^2.0*pi*h - 
                                 ((predictors[['cbh200']]/(100.0*2.0*pi)-0.01)^2.0*pi*h),
                          V250 = (predictors[['cbh250']]/(100.0*2.0*pi))^2.0*pi*h - 
                                 ((predictors[['cbh250']]/(100.0*2.0*pi)-0.01)^2.0*pi*h))

# Determine the mass of each stem section (Msection, kg)
predictors <- add_column (predictors,
                          M50  = predictors [['V50']]  * rho_w,
                          M100 = predictors [['V100']] * rho_w,
                          M150 = predictors [['V150']] * rho_w,
                          M200 = predictors [['V200']] * rho_w,
                          M250 = predictors [['V250']] * rho_w)
```

``` {r getSugarConcentration}
# Read the sugar and starch concentration (means for needles, branches and roots and first centimeter for wood sections)
suppressMessages (source ('processNSCDataForExp2017.R'))

# Create responseVariables tibble
responseVariables <- tibble (year = 2017,
                             month    = c (rep (7, 41), rep (8, 41), 
                                           rep (10, 41), rep (11, 41)),
                             tree = rep (1:41, 4),
                             sugarW250 = NA, starchW250 = NA, sugarW200 = NA,	
                             starchW200 = NA,	sugarW150 = NA,	starchW150 = NA,	
                             sugarW100 = NA,	starchW100 = NA, sugarW50 = NA,	
                             starchW50 = NA)

# Wrangle data to get sugar and starch concentrations (g kg-1) for each section and 
# multiply the concentration by the dry mass of the section (kg) to obtain total sugar and 
# starch reserves (kg)
for (r in 1:dim (responseVariables) [1]) {
  treeID <- responseVariables [['tree']] [r]
  iMonth <- responseVariables [['month']] [r]
  if (unique (stemData [['treatment']] [stemData [['treeID']] == treeID]) == 1) {
    responseVariables [['sugarW150']] [r] <- 
      predictors [['M150']] [predictors [['tree']] == treeID] * 
      (stemData [['sugar']] [stemData [['treeID']] == treeID & 
                             month (stemData [['date']]) == iMonth] / 100) * 1000
  } else if (unique (stemData [['treatment']] [stemData [['treeID']] == responseVariables [['tree']] [r]]) == 2 | unique (stemData [['treatment']] [stemData [['treeID']] == responseVariables [['tree']] [r]]) == 3) {
    responseVariables [['sugarW100']] [r] <- 
      predictors [['M100']] [predictors [['tree']] == treeID] * 
      (stemData [['sugar']] [stemData [['treeID']] == treeID & 
                             stemData [['sampleHeight']] == 1 &
                             month (stemData [['date']]) == iMonth] / 100) * 1000
    responseVariables [['sugarW200']] [r] <- 
      predictors [['M200']] [predictors [['tree']] == treeID] * 
      (stemData [['sugar']] [stemData [['treeID']] == treeID & 
                             stemData [['sampleHeight']] == 2 &
                             month (stemData [['date']]) == iMonth] / 100) * 1000
  } else if (unique (stemData [['treatment']] [stemData [['treeID']] == responseVariables [['tree']] [r]]) == 4) {
    responseVariables [['sugarW50']] [r] <- 
      predictors [['M50']] [predictors [['tree']] == treeID] * 
      (stemData [['sugar']] [stemData [['treeID']] == treeID & 
                             stemData [['sampleHeight']] == 0.5 &
                             month (stemData [['date']]) == iMonth] / 100) * 1000
    responseVariables [['sugarW150']] [r] <- 
      predictors [['M150']] [predictors [['tree']] == treeID] * 
      (stemData [['sugar']] [stemData [['treeID']] == treeID & 
                             stemData [['sampleHeight']] == 1.5 &
                             month (stemData [['date']]) == iMonth] / 100) * 1000
    responseVariables [['sugarW250']] [r] <- 
      predictors [['M250']] [predictors [['tree']] == treeID] * 
      (stemData [['sugar']] [stemData [['treeID']] == treeID & 
                             stemData [['sampleHeight']] == 2.5 &
                             month (stemData [['date']]) == iMonth] / 100) * 1000
  }
}

```

## Total above ground biomass 

According to @jenkins_comprehensive_2004, we calcualted total aboveground biomass ($B; in \; kg$) from the diameter at breast height of tree $t$ ($d_{t}; in \; m$) as follows:

$$
B = exp (\beta_0 + \beta_1 \; ln \; d_{t})
$$

``` {r totalBiomass}
biomass04 <- exp (2.4800 + 2.4835 * log (predictors [['cbh150']] / pi))  
#biomass08 <- exp (-2.6177 + 2.4638 * log (predictors [['cbh150']] / pi)) # with specific gravity < 0.45
#biomass08 <- exp (-3.0506 + 2.6465 * log (predictors [['cbh150']] / pi)) # with specific gravity >= 0.45
#biomassxx <- 2.2874 * exp (0.1968 * (predictors [['cbh150']] / pi)) # From peichl
```

with generic pine species (Pinus spp.) parameters of $\beta_0 = -2.4800$ and $\beta_1 = 2.4835$. This total biomass includes foliage, stemwood, bark and branches. To get tissue specific values for the mass of stemwood, bark and foliage we estimated the fraction of the total aboveground biomass in each tissue using parameters from Table 2 in @jenkins_comprehensive_2004 (i.e. for stemwood $\alpha_0 = –0.3737$ and $\alpha_1 = –1.8055$) and the following equation:

$$
f = exp (\alpha_0 + \frac {\alpha_1} {d_t})
$$
``` {r fractionalBiomass}
rootFraction <- exp (-1.5619 + ( 0.6614 / (predictors [['cbh150']] / pi)))
stemFraction <- exp (-0.3737 + (-1.8055 / (predictors [['cbh150']] / pi)))
barkFraction <- exp (-2.0980 + (-1.1432 / (predictors [['cbh150']] / pi)))
foliFraction <- exp (-2.9584 + ( 4.4766 / (predictors [['cbh150']] / pi)))
branFraction <- 1.0 - foliFraction - stemFraction - barkFraction

rootMass <- biomass04 * rootFraction
stemMass <- biomass04 * stemFraction
barkMass <- biomass04 * barkFraction
branMass <- biomass04 * branFraction
foliMass <- biomass04 * foliFraction
```

Branch biomass was calculated as the residual of total aboveground biomass and the three other tissues estimated above. Here, the mean stem, bark, branch and foliage fractions were `r round (mean (stemFraction) * 100.0,1)`%, `r round (mean (barkFraction) * 100.0,1)`%, `r round (mean (branFraction) * 100.0,1)`%, and  `r round (mean (foliFraction) * 100.0,1)`%, respectively.

$$
f_{branches} = 1.0 - f_{foliage} - f_{stemwood} - f_{bark}
$$
Root biomass was also estimated as a fraction of total aboveground biomass with an average of `r round (mean (rootFraction) * 100.0,1)`%.

For each organ and the entire tree we multiplied the respecitve biomass with the average sugar and starch concentration (in $% dry weight$) derived from colourimetric assays. The total nonstructural carbon estimates are the sum of the mass of sugar and starch. 

```{r totalFoliage reserves}
# Determine tissue sugar reserves
#sugarRoot   <- rootData [['sugar']] * rootMass # TR still missing values
sugarNeedle <- leafData [['sugar']] * foliMass

# Determine tissue starch reserves
#starchRoot   <- rootData [['starch']] * rootMass # TR still missing values
starchNeedle <- leafData [['starch']] * foliMass

# TR I could also calculate sugar in branches and bark in the future with the samples we still have at NAU. Furthermore, I could average the stem section of the control to derive stem values.
```

# References